---
title: "<center>GPS Processing Manual – Wildlife and Applied Genomics Lab<center>"
author: "<center>JFW<center>"
date: "<center>September 11, 2020<center>"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
theme: cerulean
---
<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkGreen;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
body {
text-align: justify
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```
# This is a general manual for dealing with and filtering/cleaning GPS data from wildlife GPS collars

### Title: GPS data filtering/cleaning
### Purpose: Filter and clean GPS data for downstream analysis
### Author: JFW
### Input: .csvs of raw data
### Date: September 11 2020
### Output: single .csv file with all individual data files cleaned and ready for downstream analysis, seasonal data csvs, figure of where movement rates exceed certain value

## General data filtering and cleaning
### Environment setup
```{r Environment setup, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
libs <- c('data.table', 'ggplot2', 'dplyr', 'readr', 'rgdal')
lapply(libs, require, character.only = TRUE)
```
### Read in all .csv files (need to have working directory set to where all the raw files are), combine, and read back in
```{r Combine all .csvs into one, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
allgps <- list.files(full.names = TRUE) %>% 
  lapply(read_csv) %>% 
  bind_rows 
# Export to new csv
#write.csv(allgps, combinedgps.csv)
#Read in exported file, encoding it with Latin1 fixes a gsub error I kept getting
allgps <- read_csv("output/combinedGPS.csv", locale = locale(encoding = "Latin1"))
```
### Filter combined csv to only contain relevant information (will be different for everyone) and set proj4string, which tells R where your animals are located according to a UTM zone
```{r Filter combined csv to only contain relevant information, echo = TRUE, warning = FALSE, eval = TRUE}
#Select only relevant columns of data
gps_simp <-allgps %>%
  select(CollarID, `Mort. Status` ,LMT_Date, LMT_Time, `Height [m]`, DOP,  `Latitude [°]`,`Longitude [°]`,
             FixType, Easting, Northing, AnimalID)
# Setting object to utm zone 9N - THIS DEPENDS ON WHERE THE DATA COMES FROM - CHECK FOR YOUR UTM ZONE
utm9N <-'+proj=utm +zone=9 ellps=WGS84'
# Renaming columns for simplicity
gps_simp <- gps_simp %>%
  rename(Height = `Height [m]`)%>%
  rename (mortality = `Mort. Status`)%>%
  rename (latitude = `Latitude [°]`)%>%
  rename (longitude = `Longitude [°]`)
```

### Filter to remove any lat/longs that are N/A, remove any mortality signals, create a minimum/maximum height value, and ensure filter for whatever date range you're looking at. This will vary on the individual dataset and objectives as well
``` {r Filter further, echo = TRUE, warning = FALSE, eval = TRUE}
# Histogram of altitude before filtering
hist(gps_simp$Height)
# Filtering data for erroneous points, mortality signals, and height restrictions
gps_simp_clean <- gps_simp %>%
  filter (latitude!="NA") %>%
  filter (longitude!= "NA")%>%
  filter(mortality=="normal")%>%
  filter(Height > 600, `Height` <2500)%>%
  filter(LMT_Date >= as.Date("2018-01-01") & LMT_Date <= as.Date("2020-10-31"))
# Histogram of altitude after filtering
hist(gps_simp_clean$Height)
# Export filtered data file
#write.csv(gps_simp_clean, "output/GPS_Simplified_Filtered.csv")
```
### Create a date time variable, and set boundaries
``` {r Create a date time variable, and set boundaries, message = FALSE, echo = TRUE, warning = FALSE, eval = TRUE}
# paste date and time to new variable 
datetime <- (paste0 (gps_simp_clean$LMT_Date, sep = " ", gps_simp_clean$LMT_Time))
# bind DateTime to data frame
gps_simp_clean$datetime <- datetime
# making data a data table
gps_simp_clean_dt<-as.data.table(gps_simp_clean)
# Set variable LMTdatetime as proper format to use further
gps_simp_clean_dt [,LMTdatetime := as.POSIXct(datetime)]
# changing coordinates and setting boundaries 
# This will create an Easting and Northing column depending on your Longitude/Latitude
gps_simp_clean_dt[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(longitude, latitude), utm9N))]
# What do our easting/northing points look like?
plot(gps_simp_clean_dt$EASTING, gps_simp_clean_dt$NORTHING,
     xaxt='n', yaxt='n', ann=FALSE)
# Generate some sort of boundaries if you see obviously erroneous points in plot above - values below are just random, boundaries depend on individual datasets
lowEast <- 100; highEast <- 100
lowNorth <- 4000; highNorth <- 5000
# Filter by bounds - This will again depend on your study extent but will help
# get rid of those random points off in space
gps_simp_clean_dt <- gps_simp_clean_dt[(lowEast < EASTING & EASTING < highEast) &
                                     (lowNorth < NORTHING & EASTING < highNorth)]
# check to see what it looks like (I removed axes as location data is sensitive)
plot(gps_simp_clean_dt$EASTING, gps_simp_clean_dt$NORTHING,
     xaxt='n', yaxt='n', ann=FALSE)
```




### Calculate number of GPS fixes per individual
``` {r Calculate number of GPS fixes per individual, message = FALSE, echo = TRUE, warning = FALSE, eval = TRUE}
#Produces tibble with animal ID and number of fixes
#Animalid is unique ID identifying individual animal
fixes_per_ind <- group_by(gps_simp_clean_dt) %>%
  group_by(AnimalID) %>%
  summarise(number_of_fixes = n())
#write.csv(fixes_per_ind, "output/NumberofFixesPerInd.csv")
```

### Calculate step length to determine distance between fixes, time between fixes, and ultimately, a movement rate
``` {r Calculate step length to determine distance between fixes, time between fixes, and ultimately, a movement rate, message = FALSE, echo = TRUE, warning = FALSE, eval = TRUE}
# data.table Step Length 
# This next part will help you produce a simple step length and then a movement rate so that you can 
# then remove any biologically impossible movements (i.e., 30km/hr movement rates) 
#Sort Data
gps_simp_clean_dt <- gps_simp_clean_dt[order(gps_simp_clean_dt$AnimalID, 
                                         gps_simp_clean_dt$datetime),]
# Set columns
time.col <- 'LMTdatetime'
coord.cols <- c('EASTING', 'NORTHING')
# Create lag and dif column names
lag.cols <- paste('lag', coord.cols, sep = '')
difference.cols <- c('difX', 'difY')
lag.time.col <- paste0('lag', time.col)
dif.time.col <- paste0('dif', time.col)
# Use shift  to create lagged cols
gps_simp_clean_dt[order(get(time.col)), (lag.cols) := shift(.SD, 1, NA, 'lag'),
                by = .(CollarID),
                .SDcols = coord.cols]
# Find the difference squared between all points in each x,y separately
gps_simp_clean_dt[, (difference.cols) := .((get(coord.cols[1]) - get(lag.cols[1])) ^2,
                                         (get(coord.cols[2]) - get(lag.cols[2])) ^2)]
# Square root the summed difference for a simple step length
gps_simp_clean_dt[, simpleStep := sqrt(rowSums(.SD)),
                .SDcols = difference.cols]
## Delta Time
gps_simp_clean_dt[order(get(time.col)), (lag.time.col) := shift(.SD, 1, NA, 'lag'), 
                by = .(CollarID),
                .SDcols = time.col]
# difference in time in hours
gps_simp_clean_dt[, (dif.time.col) := as.numeric(get(time.col) - get(lag.time.col), units = 'hours')]
# Simple step length divided by time difference - something here is generating infinites
gps_simp_clean_dt[, moveRate := simpleStep / (get(dif.time.col))]
# Output pdf of locs xy BEFORE movement rate cutoff and flag the locs----
# where moverate is higher than 15 km/h. If you simply want to view the plots
# and not save them you can just run the lines in between pdf and dev.off()
#pdf('graphics/flaglocs30kmh.pdf')
moverateflag<- ggplot(gps_simp_clean_dt) +
  geom_point(aes(EASTING,NORTHING)) + 
  geom_point(aes(EASTING, NORTHING), color = 'red', data = gps_simp_clean_dt[moveRate > 30000]) + 
  labs( gps_simp_clean_dt= 'EASTING', y = 'NORTHING') + guides(color = FALSE)+
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
moverateflag
#dev.off()
# Drop more than 15km/hr movements
gps_simp_clean2_dt <- gps_simp_clean_dt%>%
  filter(moveRate<30000)
# Export it 
#write.csv(gps_simp_clean2_dt, 'output/gps_FilteredCleaned.csv')
```
## Seasonal data preparation - these dates correspond to my mountain goat data, will change depending on study system
### Setting up seasonal data sets 

``` {r Seasonal data prepration, message = FALSE, echo = TRUE, warning = FALSE, eval = TRUE}
winter2018 <- gps_simp_clean2_dt %>%
  filter(LMT_Date >= as.Date("2017-12-01") & LMT_Date <= as.Date("2018-04-30"))
winter2019 <- gps_simp_clean2_dt %>%
  filter(LMT_Date >= as.Date("2018-12-01") & LMT_Date <= as.Date("2019-04-30"))
winter2020<- gps_simp_clean2_dt %>%
  filter (LMT_Date >=as.Date("2019-12-01") & LMT_Date <= as.Date("2020-04-30"))
winter <- rbind(winter2018, winter2019, winter2020)
#write.csv(winter, "output/winterGPS.csv")
#### Setting up seasonal data sets - SUMMER 
summer2018 <- gps_simp_clean2_dt %>%
  filter(LMT_Date >= as.Date("2018-05-01") & LMT_Date <= as.Date("2018-10-31"))
summer2019 <- gps_simp_clean2_dt %>%
  filter(LMT_Date >= as.Date("2019-05-01") & LMT_Date <= as.Date("2019-10-31"))
summer2020 <- gps_simp_clean2_dt%>%
  filter (LMT_Date >=as.Date("2020-05-01") & LMT_Date <= as.Date("2020-10-31"))
summer <- rbind(summer2018, summer2019, summer2020)
#write.csv(summer, "output/summerGPS.csv")